### 2013-01-05 ISMB/methods_ISMB.tex
algo

\newcommand{\PE}[1]{E(#1)}
\newcommand{\EI}{\text{EI}}
\newcommand{\ES}{\text{ES}}
\newcommand{\ISO}{\text{ISO}}
We introduce a probabilistic model, aiming to design RNA sequences with a 
specific $\Gb+\Cb$ content, folding into a predefined 2D conformation.
We use a Boltzmann weighted distribution based on a pseudo-energy function 
$\PE{\cdot}$ taking into accounts stacked canonical base pairs free-energy.

%To that purpose, a Boltzmann weighted distribution is used, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both the free-energy and its putative isostericity towards a multiple sequence alignment. In this model, the probability that the nucleotide at a given position needs to be mutated (i.e. corresponds to a sequencing error) can be computed using a variant of the \emph{Inside-Outside algorithm}~\cite{Lari1990}.
%\subsection{Probabilistic model}
%Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
%then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
%  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
%where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
%are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.
%\subsubsection{Energy contribution}
%The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
%  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
%where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).

@@

%\subsubsection{Definitions}

@@

%  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
%\subsubsection{Inside}
%\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
%\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
%can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
%(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
%or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
%In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
%and locally instanciated bases.}
%The \emph{Inside} function $\Z{i,j}{m}{a,b}$ is the partition function, i.e. the sum of Boltzmann factors, over all sequences in the interval $[i,j]$, at distance $m$ of $s_{[i,j]}$, and having flanking nucleotides $a$ and $b$ (at positions $i-1$ and $j+1$ respectively). 
%Such terms can be defined by recurrence, for which the following initial conditions holds:
%	\forall i \in [0,n-1]:\, \Z{i+1,i}{m}{a,b}=\left\{
%		1 &\text{If } m = 0\\
%		0 &\text{Otherwise.}
%\label{eq:Z_in}
%In other words, the set of sequences at distance $m$ of the empty sequence is either empty if $m>0$, or restricted to the empty sequence, having energy $0$, if $m=0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions.

The main recursion is composed of four terms:
\begin{equation}
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}
			 \cdot \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
      \cdot\Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \cdot\Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Otherwise}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation}
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is unpaired, then 
any sequence consists in a, possibly mutated, nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ having either $m-\Kron_{a',s_i}$, accounting for a possible mutation 
at position $i$, and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two, possibly mutated, nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$ having distance $m-\Kron_{a'b',s_is_j}$ (to avoid exceeding the targeted distance $m$).
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$. Since the number of mutations sum to $m$ over the whole sequence must , then a parameter $m'$ is introduced to distribute the remaining mutations between the two sequences.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

>>>>

\newcommand{\PE}[1]{E(#1)}
\newcommand{\EI}{\text{EI}}
\newcommand{\ES}{\text{ES}}
\newcommand{\ISO}{\text{ISO}}
\newcommand{\Prob}{\text{P}}

We introduce a probabilistic model, aiming to design RNA sequences with a 
specific $\Gb+\Cb$ content, folding into a predefined 2D conformation.
We use a Boltzmann weighted distribution based on a pseudo-energy function 
$\ES(\cdot)$ taking into accounts stacked canonical base pairs free-energy.

%To that purpose, a Boltzmann weighted distribution is used, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both the free-energy and its putative isostericity towards a multiple sequence alignment. In this model, the probability that the nucleotide at a given position needs to be mutated (i.e. corresponds to a sequencing error) can be computed using a variant of the \emph{Inside-Outside algorithm}~\cite{Lari1990}.
%
%\subsection{Probabilistic model}
%Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
%then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
%\begin{align*}
%  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
%\end{align*}
%where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
%are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.
%
%\subsubsection{Energy contribution}
\subsection{Definitions}
\subsubsection{Profile}
At every position, the nucleotide distribution is the same. 
It is first initialized equiprobably. A bisection scheme is then applied until the sampled sequences $\Cb+\Gb$ content is as desired. The probability
of a given nucleotide $a\in\B$ will be given by $\Prob(a)$ at every 
position.

\subsubsection{Energetic model}

The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by

\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}

where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).

@@

@@

%\begin{equation}
%  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
%\end{equation}
%
%
%\subsubsection{Inside}
%\begin{figure}
%\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
%\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
%can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
%(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
%or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
%In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
%and locally instanciated bases.}
%\end{figure}
%
%The \emph{Inside} function $\Z{i,j}{m}{a,b}$ is the partition function, i.e. the sum of Boltzmann factors, over all sequences in the interval $[i,j]$, at distance $m$ of $s_{[i,j]}$, and having flanking nucleotides $a$ and $b$ (at positions $i-1$ and $j+1$ respectively). 
%Such terms can be defined by recurrence, for which the following initial conditions holds:
%
%\begin{equation}
%	\forall i \in [0,n-1]:\, \Z{i+1,i}{m}{a,b}=\left\{
%	\begin{array}{ll}
%		1 &\text{If } m = 0\\
%		0 &\text{Otherwise.}
%	\end{array}\right.
%\label{eq:Z_in}
%\end{equation}
%In other words, the set of sequences at distance $m$ of the empty sequence is either empty if $m>0$, or restricted to the empty sequence, having energy $0$, if $m=0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions.

The main recursion is composed of four terms:
\begin{equation}
	\Z{i,j}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{a'\in \B}  
      \Prob(a')
      \cdot\Z{i+1,j}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{a',b'\in \B^2}
			 \Prob(a')
			 \cdot\Prob(b')
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{i+1,j-1}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{a',b'\in \B^2}
      \Prob(a')
			\cdot\Prob(b')
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{i+1,k-1}{a',b'}
      \cdot\Z{k+1,j}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Otherwise}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation}
The cases can be broken down as follows:
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is unpaired, then 
any sequence consists in a nucleotide $a'$ at position  $i$, 
followed by a sequence over $[i+1,j]$ and having flanking nucleotides $a'$ and $b$.
\item[$S_i=j$ and $S_{i-1}=j+1$:] Any sequence generated in $[i,j]$ consists of two nucleotides $a'$ and $b'$, flanking a sequence over $[i+1,j-1]$.
Since positions $i$ and $i-1$ are paired with $j$ and $j+1$ respectively, 
then a stacking energy contribution is added. 
\item[$S_i=k$ and $i<k \leq j$:] If position $i$ is paired and not involved in a stacking, then the 
only term contributing directly to the energy is the nucleotide distribution $\Prob(\cdot)$. 
Any sequence on $[i,j]$ consists of two nucleotide $a'$ and $b'$ at positions $i$ and $k$ respectively, flanking a sequence over interval $[i+1,k-1]$ and preceding a (possibly empty) sequence interval $[k+1,j]$.
\item[Else:] In any other case, we are in a derivation of the SCFG that does not correspond to the secondary structure $S$, and we return $0$.
\end{description}

\subsection{Stochastic backtrack}

\begin{algorithm}
	\SetAlgoLined
	\lIf{$i>j$} \Return{$\varepsilon$};\tcp*[f]{Empty sequence}\\
	rand $\leftarrow$ Random$\left(\Z{i,j}{a,b}\right)$\;
	k $\leftarrow S_i$\;
	\If(\tcp*[f]{Unpaired}){$k = -1$}{
		\For{$a'\in\B$}{
			rand $\leftarrow$ rand $- \Prob(a')\cdot \Z{i+1,j}{a',b}$\;
			\lIf{$\text{rand}<0$}\Return{$a'\cdot\text{Backtrack}\left((i+1,j),(a',b),S\right)$}\;
		}
	}
	\Else{
		\If(\tcp*[f]{Stack}){$k=j \land S_{i-1}=j+1$}{
			\For{$(a',b')\in\B\times\B$}{
				rand $\leftarrow$ rand $-
			 	\Prob(a')
			 	\cdot\Prob(b')
			 	\cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 	\cdot \Z{i+1,j-1}{a',b'}	$\;
				\If{$\text{rand}<0$}{
					\Return{$a'\cdot\text{Backtrack}\left((i+1,j-1),(a',b'),S\right)\cdot b'$}\;		
				}
			}
		}
		\Else(\tcp*[f]{Paired not stacked}){
			\For{$(a',b')\in\B\times\B$}{
				rand $\leftarrow$ rand $-
				\Prob(a')
				\cdot\Prob(b')
				\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    \cdot\Z{i+1,k-1}{a',b'}
    	  \cdot\Z{k+1,j}{b',b}$\;	
 				\If{$\text{rand}<0$}{
					\Return{$a'
					\cdot\text{Backtrack}\left((i+1,k-1),(a',b'),S\right)
					\cdot b'
					\cdot\text{Backtrack}\left((k+1,j),(b',b),S\right)
					$}\;	
				}
			}
		}
	}
\caption{Backtrack$\left((i,j),(a,b),S\right)$\label{alg:back}}
\end{algorithm}
